---
title: "BD^2^ second pilot barcode demultiplexing and basic read statistics"
author: Osama Arshad
date: last-modified
format: 
  html:
    toc: true
    toc-location: right-body
    toc-depth: 4
    embed-resources: true
editor: source
echo: false
---

```{r}
#| label: load packages
#| include: false

library(tidyverse)
library(knitr)
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(scales)
library(ggstats)
library(readxl)
library(janitor)
library(here)
library(plotly)
library(htmlwidgets)
library(kableExtra)
library(DT)
```

```{r}
#| label: read sample sheet
samplesheet <- read_excel(here("data/Pilot2_metadata_cleaned.xlsx"), n_max = 40) |>
  as_tibble(.name_repair = janitor::make_clean_names) |>
  # rename columns
  dplyr::rename(protocol = "protocol_deviation") |>
  # call the protocol where there is no deviation the "original" protocol
  dplyr::mutate(protocol = ifelse(protocol=="None", "Original", protocol)) |>
  mutate(protocol = fct_relevel(protocol, c("Original", "Biotinylated RT", "Size Selected"))) |>
  # change barcode number to labels e.g. 2 -> barcode02 to be able to match with stats table
  mutate(barcode = paste0("barcode", sprintf("%02d", as.numeric(barcode)))) |>
  # add prefix "pool" to columns where capture_pool is numeric (original protocol) e.g. 1 -> Pool 1
  mutate(capture_pool = ifelse(str_detect(capture_pool, "^[:digit:]{1}$"), str_c("Pool ", capture_pool), capture_pool))
```

```{r}
#| label: read demux stats
demux_stats <- read.table("results/seqkit_stats/barcode_classification_table.txt", sep = "\t", header = TRUE) |>
  # extract the barcode part of filename for all files except the "unclassified" fastq 
  mutate(barcode = ifelse(str_detect(file, "unclassified"), "unclassified", str_extract(file, regex("barcode\\d{2}"))))
  
demux_stats <- left_join(demux_stats, samplesheet, by = "barcode") |>
  # create a column that captures whether barcode is assigned to a sample or not
  mutate(class = ifelse(barcode %in% samplesheet$barcode, "sample", "unused"))
```

## QC passed sequencing output

-   Reads: `{r} round(sum(demux_stats$num_seqs)/1E6, 1)` Million.
-   Throughput: `{r} round(sum(demux_stats$sum_len)/1E9, 1)` Gigabases.

## Barcode Classification Summary

```{r}
#| label: classification rate

# total number of reads
n_reads <- sum(demux_stats$num_seqs)
# number classified 
n_classified <- sum(demux_stats$num_seqs[str_detect(demux_stats$barcode, "barcode")])
n_unassigned <- sum(demux_stats$num_seqs[demux_stats$barcode=="unclassified"])

plot_df <- data.frame(label = c("Classified", "Unclassified"), n = c(n_classified, n_unassigned)) |>
  mutate(label = fct_relevel(label, c("Unclassified", "Classified"))) |>
  mutate(n = n/1E6) |>
  mutate(dummy = "dummy")

# https://cran.r-project.org/web/packages/ggstats/vignettes/stat_prop.html
plot_df |> 
  ggplot() +
  aes(x = dummy, fill = label, weight = n) +
  geom_bar() +
  geom_text(
    aes(label = scales::percent(after_stat(prop), accuracy = 0.01)),
    stat = "prop",
    position = position_stack(.5)
  ) +
  labs(y = "No. of Reads (Mn)") +
  scale_fill_manual(values=c("#999999", "#E69F00")) +
  theme_minimal() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        legend.title = element_blank()
  )
```

From the plot above:

-   Total input QC passed basecalled reads: `{r} round(n_reads/1E6, 1)` Million.
-   Classified (assigned to a barcode): `{r} round(n_classified/1E6, 1)` Million.

resulting in a barcode classification rate of `{r} round((n_classified/n_reads)*100, 1)`%.

```{r}
# remove the unassigned stats from dataframe for analyses to follow
demux_stats <- demux_stats[demux_stats$barcode != "unclassified", ]
  
# barcodes with number of sequences below this threshold are considered
# background, others high count
background_cutoff <- 10000

demux_stats <- demux_stats |>
mutate(count_category = ifelse(num_seqs <= background_cutoff, "background", "regular")) |>
  mutate(count_category = fct_relevel(count_category, "regular", "background")) |>
  # create a column called 'percent' which represents fraction of reads
  # of total column contains
  # do not multiply by 100 as DT package will do so when displaying table 
  mutate(percent = num_seqs/sum(num_seqs))
```


## Read assignment by barcode

The distribution of reads detected for each barcode is shown below. Barcodes with less than `{r} background_cutoff/1000`k reads are considered background, rest *"regular"*.

```{r}
#| fig-height: 11

demux_stats |>
  mutate(num_seqs = num_seqs/1000) |>
  ggplot(
    aes(
      x = reorder(barcode, -num_seqs), 
      y = num_seqs, 
      fill = class)
    ) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values=c("#E69F00", "#999999")) +
  labs(y = "No. of Reads (K)", fill = "barcode status") +
  facet_wrap(~count_category, scales = "free", ncol = 1) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom", 
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_blank(),
  )
```

```{r}
# remove background barcodes, unassigned barcodes with no metadata
demux_stats <- demux_stats |>
  filter(count_category != "background", class == "sample")
```


```{r}
# calculate median no. of reads for sample barcodes
median_nreads_stats <- demux_stats |>
  mutate(num_seqs = num_seqs/1000) |>
  summarise(min = min(num_seqs), median = median(num_seqs), max = max(num_seqs))
```

-   Unassigned barcodes have hardly any counts.
-   All sample barcodes have at least `{r} prettyNum(formatC(median_nreads_stats$min, format = "d"), big.mark = ",")`k reads assigned to them.
-   Sample barcodes have median of `{r} prettyNum(formatC(median_nreads_stats$median, format = "d"), big.mark = ",")`k reads.
-   Assuming a uniform distribution of reads across barcodes, given that we have `{r} round(n_classified/1E6, 1)` Million classified reads and 40     samples, we roughly expect each barcode to have `{r} round((n_classified/40)*(1/1E6), 2)` Mn reads.

### Number of reads annotated by protocol

```{r}

# function to plot barcode of counts by metadata column
plot_counts <- function(demux_stats, metadata_column) {
  stopifnot(metadata_column %in% colnames(demux_stats))
  
  demux_stats |>
  mutate(num_seqs = num_seqs/1000) |>
  ggplot(
    aes(
      x = reorder(barcode, -num_seqs), 
      y = num_seqs, 
      fill = !!sym(metadata_column))
    ) +
  geom_bar(stat = "identity") +
  labs(y = "No. of Reads (K)") +
  theme_minimal() +
  theme(
    legend.position = "bottom", 
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_blank(),
  )
}
```

```{r}
plot_counts(demux_stats, "protocol")
```

- 8 of the 9 barcodes with less than 250K reads are size selected (but so are the two with the most reads)

### Number of reads by protocol (and brain region)

```{r}
demux_stats |>
  ggplot(aes(x = protocol, y = num_seqs, color = tissue_origin)) +
    geom_jitter(position = position_jitter(seed = 1)) +
    labs(y = "Number of Reads (k)") + 
    theme_bw()
```

## N50 by barcode

Read length (N50) distribution for barcodes:

```{r}
demux_stats |>
  ggplot(
    aes(
      x = reorder(barcode, -N50), 
      y = N50
      )
    ) +
  geom_bar(stat = "identity") +
  scale_y_continuous(name="N50", labels = scales::comma) +
  theme_minimal() +
  theme(
    legend.position = "none", 
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_blank(),
  )
```

-   Min: `{r} prettyNum(summarise(demux_stats, min(N50)), big.mark = ",", format = "d")`k
-   Max: `{r} prettyNum(summarise(demux_stats, max(N50)), big.mark = ",", format = "d")`k
-   Median: `{r} prettyNum(summarise(demux_stats, median(N50)), big.mark = ",", format = "d")`k

### N50 annotated by protocol

```{r}
demux_stats |>
  ggplot(
    aes(
      x = reorder(barcode, -N50), 
      y = N50,
      fill = protocol
      )
    ) +
  geom_bar(stat = "identity") +
  scale_y_continuous(name="N50", labels = scales::comma) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_blank(),
  )
```


### N50 by protocol (and brain region)

```{r}
demux_stats |>
  ggplot(aes(x = protocol, y = N50, color = tissue_origin)) +
    geom_jitter(position = position_jitter(seed = 1)) +
    labs(y = "N50") + 
    theme_bw()
```

Median N50 by barcode for each protocol is:

```{r}
# calculate median N50 by protocol

# threshold to seperate into upper and lower bins
demux_stats |>
  group_by(protocol) |>
  summarise(median = median(N50)) |>
  kbl() %>%
  kable_styling()
```

## Data

```{r}
demux_stats %>%
  dplyr::select(barcode, num_seqs, N50, everything()) %>%
  dplyr::select(-file, -format, -type, -class, -percent) %>% 
  # https://martinctc.github.io/blog/vignette-downloadable-tables-in-rmarkdown-with-the-dt-package/
  DT::datatable(extensions = 'Buttons',
                options = list(dom = 'Blfrtip',
                           buttons = c('copy', 'csv', 'excel'),
                           lengthMenu = list(c(10,25,50,-1),
                                             c(10,25,50,"All")))
                ) %>% 
  formatCurrency('num_seqs', currency = "", interval = 3, mark = ",", digits = 0) %>% 
  formatCurrency('N50', currency = "", interval = 3, mark = ",", digits = 0)
```
